Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
Shortenings so far.

@@

To enable multitasking through cooperative kernels, the runtime (in
our case, the mega-kernel), must (a) track the state of all
workgroups, i.e.\ if a workgroup is waiting or computing a kernel (b)
maintain consistent context states for each kernel, i.e.\ the number
of active workgroups computing a cooperative kernel and (c) provide a
safe way for these states to be modified (according to the semantics
of $\offerfork$ and $\offerkill$. We discuss how that is achieved
here.

@@

When a workgroup executes a kernel (cooperative or otherwise), it must
be able to query a consistent context to obtain a workgroup id value
(starting from zero up to the number of executing workgroups) and the
number of workgroups executing the kernel. In single kernel execution,
the OpenCL primitives {\tt get\_num\_groups()} and {\tt
  get\_group\_id()} provide this functionality natively. For multiple
kernel execution, we must emulate these primitives, per kernel,
explicitly through \emph{kernel contexts}. This structure contains:
(a) the number of active groups, (b) the total number of groups asked
for, and (c) a mapping from the native workgroup id to a new workgroup
id valid for a particular kernel. Our framework requires the native
kernel context (to obtain a global base workgroup id) as well as an
additional distinct kernel context for each kernel running
concurrently. \TSComment{a driver implementation would likely generate
these on the fly}.

@@

In order to dynamically manage workgroups executing cooperative
kernels, our framework must (a) track the state of each workgroup and
(b) provide a channel of communication from the scheduler to
workgroups executing $\offerfork$ and $\offerkill$. To achieve this,
we use a \emph{Scheduler context} structure which contains a mapping
from primitive workgroup ids to their state i.e.\ waiting or computing
a kernel as well as an integer value describing the number of
workgroups in each state. The scheduler can then send cooperative
kernels a \emph{resource message}, indicating if workgroups executing
the cooperative kernel need to (a) exit at $\offerkill$ or (b)
transmit their state to forking workgroups at $\offerfork$. Thus, the
scheduler context needs a communication channel for each cooperative
kernel. In our case, the communication channel is implemented using
global atomic variables and OpenCL atomic operations to ensure
consistency.

@@

As in native OpenCL, the host maintains the ability to launch kernels.
In our framework, the host has the additional ability to request that
a kernel be cooperative \emph{and} to specify the amount of
\emph{resource} each kernel requires. In our implementation, the only
resource we have explicit control over is the number of concurrently
executing workgroups; implementations closer to metal might also
consider local memory and registers as resource.

To launch a kernel, the host sends a data packet to the GPU, which
consists of: a kernel to execute, the inputs to the kernel, the amount
of resource to use for the kernel (which includes the number of
workgroups to execute the kernel), and a flag indicating if the kernel
is cooperative. In our implementation, this communication channel
between host and device is achieved through fine-grained SVM atomics.
The kernel specified must be one of the kernels merged inside the
mega-kernel. If there are large constant inputs to the kernel, these
can be copied before mega-kernel launch to decrease the amount of SVM
traffic.

\paragraph{Dynamically managing workgroups}

The GPU scheduler then receives the data packet from the host
describing a kernel $K$ to be executed and now must make decisions
about how to schedule the $K$. Consider the case where $K$ requires
$W$ workgroups. The scheduler can query the scheduler context and if
there is at least $W$ available workgroups, then $K$ can be scheduled
immediately. If there are not enough available workgroups (say there
are $W_a$ workgroups available), but a cooperative kernel $K_c$ is
executing. The scheduler can signal $K_c$ through $K_c$'s
communication channel in the scheduler context communication
requesting that $K_c$ provide $W - W_a$ workgroups (via $\offerkill$)
as soon as possible. The scheduler then waits until $W$ workgroups
become available.

When $W$ workgroups become available, the scheduler initialises a
kernel context $KC$ for $K$, setting the number of active workgroups
to $W$ and mapping available workgroups (located in the scheduler
context) to their new $KC$ workgroup ids. The scheduler then sends $W$
workgroups from the available workgroups to execute kernel $K$ with
kernel context $KC$.

In the case where a cooperative kernel $K_c$ is executing and there
are $W$ avaiable workgroups, the scheduler may decide to give $K_c$
$W_k <= W$ additional workgroups via an $\offerfork$ command issued by
$K_c$. To do this, scheduler signals $K_c$ through the $K_c$'s
scheduler context communication channel to indicate the number of
workgroups that are allowed to join at the next $\offerfork$ command.
Then, when a workgroup $w$ of $K_c$ executes $\offerfork$, $w$
observes $W_k$, and takes $W_k$ workgroups from the pool of available
workgroups, provides them with transmit variables, target, and updated
kernel context.
\TSComment{this is sloppy}

>>>>

\newcommand{\KC}{\mathit{KC}}

@@

To enable multitasking through cooperative kernels, the runtime (in
our case, the mega-kernel), must (a) track the state of all
workgroups, i.e.\ if a workgroup is waiting or computing a kernel, (b)
maintain consistent context states for each kernel, e.g.\ the number
of active workgroups computing a cooperative kernel, and (c) provide a
safe way for these states to be modified in response to $\offerfork$ and $\offerkill$. We now discuss these issues in more detail.

@@

When a workgroup executes a kernel (cooperative or otherwise), it must
be able to query a consistent context to obtain a workgroup id value
and the
number of workgroups executing the kernel. In single kernel execution,
the OpenCL primitives $\getnumgroups()$ and $\getgroupid()$ provide this functionality natively. For multiple
kernel execution, we emulate these primitives via a per kernel \emph{context}, containing, for a given kernel: the number of active groups $M$, the total number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
id valid for the kernel.  In the case of a non-cooperative kernel, $M$ is undefined because the notion of active workgroups is not relevant.

@@

To dynamically manage workgroups executing cooperative
kernels, our framework must track the state of each workgroup and
provide a channel of communication from the scheduler to
workgroups executing $\offerfork$ and $\offerkill$. To achieve this,
we use a \emph{scheduler context} structure, mapping a primitive workgroup id the workgroup's status, which is either \emph{available}, or the id of the kernel that the workgroup is currently executing.  The scheduler can then send cooperative
kernels a \emph{resource message}, commanding workgroups executing
the kernel to exit at $\offerkill$, or 
spawn additional workgroups at $\offerfork$. Thus, the
scheduler context needs a communication channel for each cooperative
kernel. In our case, the communication channel is implemented using
global atomic variables.

@@

To launch a kernel, the host sends a data packet to the GPU
consists of: a kernel to execute, the inputs to the kernel, the
number of workgroups that should execute the kernel, and a flag indicating whether the kernel
is cooperative. In our implementation, this communication channel
between host and device is achieved through fine-grained SVM atomics.

\paragraph{Dynamically managing workgroups}
On receiving a data packet from the host
describing a kernel launch $K$, the GPU scheduler must make decisions
about how to schedule $K$. Suppose $K$ requests
$N$ workgroups. The scheduler queries the scheduler context.  If
there are at least $N$ available workgroups, $K$ can be scheduled
immediately. Suppose instead that there are only $N_a < N$ available workgroups, but a cooperative kernel $K_c$ is
executing. The scheduler can use $K_c$'s channel in the scheduler context to command $K_c$ to provide $N - N_a$ workgroups via $\offerkill$.
Once $N$ workgroups are available, the scheduler initialises a
kernel context $\KC$ for $K$, setting the number of active workgroups
to $N$ and mapping available workgroups (located in the scheduler
context) to their new $\KC$ workgroup ids. The scheduler then sends $N$
workgroups from the available workgroups to execute kernel $K$ with
kernel context $\KC$.  Note that if the new kernel $K$ is itself a cooperative kernel, the scheduler could decide to provide $K$ with fewer than $N$ active workgroups initially.

If a cooperative kernel $K_c$ is executing with fewer workgroups than it initially requested, the scheduler may decide to give $K_c$
additional workgroups if $K_c$ calls $\offerfork()$. To do this, the scheduler asynchronously signals $K_c$ through $K_c$'s
channel to indicate the number of
workgroups that should join at the next $\offerfork$ command.
When a workgroup $w$ of $K_c$ subsequently executes $\offerfork$, thread 0 of $w$ updates the kernel and scheduler contexts so that the given number of new workgroups are directed to the program point after the $\offerfork$ call.  This involves selecting workgroups whose status is \emph{available}, as well as copying the values of $\transmit$-annotated variables to the new workgroups.
